The Adaptive Multiscale Simulation Infrastructure -- Core Features and Biological Tissue Simulation Feature Utilization

### 2015-10-31 doc/articles/amsi_overview/amsi_overview.tex
work on a manuscript.... very rough so far

\title{}
\author{W.R. Tobin, D. Fovargue, V.W.L. Chan, M.S. Shephard}
\date{\today}
\documentclass[11pt]{article}
\bibliographystyle{siam}

@@

\label{single_scale_simulations}
Numerical simulation of physical systems is typically conducted using single scale models. The physical terms in these models are all situated within a narrow specturm of physical magnitudes surrounding the primary solution scale. The primary solution scale is determined primarily by the desired resolution of two classes of physical terms. The resoulution of the solution produced by simulation must be fine enough to be of utility to the end-user, and the resolution of variation of physical quantities required by the physical model must be fine enough to capture suitable variation to control the buildup of numerical error.

Single-scale models provide an averaged or generally accurate model of the overall evolution of a physical system in response to stimuli.

However all physical problems are multi-scale in nature, as the response of a system at the primary (or engineering) scale is in reality the result of actions taking place at scales many orders of magnitude removed.

\label{multi_scale_simulations}

Introduction of influence from scales orders of magnitude removed from the primary scale can take place in several different ways. This influence can be incorporated during the mathematical development of the physical model itself. In this case the multi-scale features are incorporated into the numerical implementation automatically, as a result of their presence in the mathematical model.

Multi-scale influences can also be incorporated in a more strictly numerical sense, combining physical models operating at different scales referencing and effecting shared physical fields and values \cite{shenoy97} \cite{weinan2003heterogenous}. Numerically multi-scale simulations associate many numerical implementations of single-scale physical models and cause them to interact in a meaningful way by passing key physical terms between scales, typically this will involve up-/down-scaling equations to take into account the difference in scale, refered to as compression and reconstruction operators in \cite{weinan2003heterogenous} for up-scaling and down-scaling, respectively.

In order for a numerically multi-scale approach to be efficacious the domain at the scale of interest (for brevity the engineering scale) must be sufficiently large that conducting the entire simulation with terms and granularity orders of magnitude smaller would make the simulation computationally intractable. Thus some computationally reasonable subset of the engineering scale is assigned tertiary scales to simulate. The precise choice of location of the tertiary simulations is largely dependent on the specific numerical algorithms used to implement the engineering scale simulation.

Typically the tertiary scale domain subset is related to locations in the engineering domain where the impact of the tertiary scales has the most impact on the engineering scale simulation.

\label{multiscale_paradigms}
Among numerically multi-scale simulations, there are two primary paradigms: concurrent and sequential multi-scale.

The sequential multi-scale paradigm involves the precomputation of parameters using a microscale simulation. These parameters are then used in a constitutive model for the macro-scale simulation \cite{garcia2008sequential}.

In the concurrent multi-scale paradigm, both scales are computed simultaneously and the required inter-scale coupling values are provided just-in-time \cite{zeng2010concurrent}.

\section{Adaptive Multiscale Simulation Infrastructure (amsi)}\label{amsi}

>>>>

\title{The Adaptive Multiscale Simulation Infrastructure -- Core Features and Biological Tissue Simulation Feature Utilization}
\author{W.R. Tobin, D. Fovargue, V.W.L. Chan, M.S. Shephard}
\date{\today}
\documentclass[11pt]{article}
\bibliographystyle{siam}

@@

\section{Biotissue Multiscale Problem}\label{biotissue}
[An overview of the biotissue problem, along with relevant references. Just needs to be sufficient for the the reader to understand the multi-scale structure of the problem, so they can follow along as specific sections of amsi are described, and then how biotissue uses those sections of amsi is discussed.]

\section{Adaptive Multiscale Simulation Infrastructure (AMSI)}\label{amsi}
The Adaptive Multi-scale Simulation Infrastructure (AMSI) a set of libraries providing support for the implementation and execution of dynamic numerically multi-scale simulations on massively-parallel HPC machines. Currently AMSI provides interfaces for describing and managing the execution of individual scales in multi-scale simulations, for defining and managing the scale-coupling communication required by such simulations, and for planning and enacting scale-sensitive load balancing operations on individual scales in a multi-scale system. AMSI is designed to ease the incorporation of legacy single-scale simulation codes into multi-scale simulations. This is accomplished by minimizing the insertion of calls to the AMSI systems in a legacy simulation code to only those locations associated with the desired multi-scale interaction.

Centralized control, while simple, is not possible in massively parallel simulation codes -- both single- and multi-scale -- as centralized dynamic control decisions require global barrier operations to effectively implement, resulting in lowered process usage and an increased time-to-solution. Decentralized control is thus required to achieve performant operation with the desired level of dynamic control flexibility. An initial approach taken to implement decentralized control decisions is a combination of deterministic control algorithms and decision implementation deferment. Deterministic algorithms operating independently on the same set of data will all arrive at the same result, and thus no further communication with respect to local implementation of such a control decision is necesarry. Decision implementation deferement is simply delaying the implementation of a specific control decision - which might otherwise introduce a parallel barier - until a suitable barier operation at which the control decision may be implemented is reached.

\section{Simulation Metamodel}\label{meta_model}
AMSI runtime operations are predicated on a carefully designed set of simulation metadata. Minimal memory and computational footprints are a primary focus in the design and implementation of this metadata, in order to curtail any impact on underlying simulation codes used in the multi-scale simulation.

The metadata retained by an individual rank is limited only to the metadata that rank will need to operate on during the execution of a simulation. There are four levels of the metadata hierarchy: rank-local, scale-local, scale-coupling, and global metadata. The determination of the minimal set of required metadata from each level of the hierarchy for an individual rank to possess during the progression of the multi-scale simulation is what concerns a large portion of the metadata management algorithms implemented in AMSI.

Local metadata currently consists primarily of simple information such as the set of scales assigned to the rank (currently limited to just a single scale, but overlapped scale assignment is a very desireable future development, see discussion in \ref{future_work}) as well as the scale-ranks of the local rank. The most important local metadata is the discrete quantity of locally-owned inter-scale data terms called a DataDistribution. The actual type(s) of the user data are unimportant to the metadata operations offered by AMSI, so long as the data can be serialized for parallel communication. This DataDistribution quantity is specific to a single scale assigned to the rank, and any number of DataDistributions can be created on each scale. The DataDistribution is converted to scale-local metadata using an AMSI process called assembly. This allows scale-local control decisions to be made using independently on each rank in a scale as opposed to globally.

Scale-local metadata is data modeling terms distributed across an entire scale, these are typically generated via assembly. Scale-local metadata such as an assembled DataDistribution are used to plan inter-scale communication, producing a piece of scale-local metadata called the CouplingPattern. The CouplingPattern models the parallel communication of coupling terms from the local scale to some coupled foreign scale. The CouplingPattern is transformed into a piece of scale-coupling metadata through a process called reconciliation. Naturally not all ranks in the foreign scale need to know the entire CouplingPattern, only the rank-local information which will guide individual parallel communication. Thus reconciliation is not communication of the entire CouplingPattern, but only individual terms of the coupling pattern, producing rank-local metadata on the foreign scale.

Another key peice of scale-local metadata is the RankSet associated with non-local but coupled scales. The MPI_Comm of the foreign scale is necessarily MPI_COMM_NULL on the local rank set (as scales are currently non-overlapping), so explicit modeling of the ranks associated with coupled foreign scales is required.  Explicit storage of individual ranks assigned to foreign (and local) scales is avoided wherever possible in order to reduce single-rank memory impact as the simulation grows.

Global metadata is currently nonexistent, maintaining zero global metadata is important to the design philosophy of AMSI, though it may be required at some point. Though global metadata determined essential to AMSI will be designed to be exteremely minimal.

\label{amsi_scales}
Individual scales in a simulation are declared during the initialization of the simulation, follow by the determination and asignment of RankSets to scales, guided by a supplied configuration file or user-implemented algorithm. Scale-coupling relations are also declared at initialization time, though this decleration is limited to the indication that two scale may interact at some point during the simulation, and nothing more.

RankSets are simply mathematical sets of MPI ranks implemented so as to take advantage of any mathematical conveniences to minimize explicit storage whenever possible.

Scales are declared to be related if some quantities of interest (such as tensor field values) are transmitted between them for scale-coupling. This coupling relationship is not assumed to be symmetric, thus if scale A is related to scale B, scale B is not implicitly related to scale A, though this may be seperately declared.

\label{amsi_communication}
Scale coupling communication is handled by the AMSI system using DataDistributions and CommuncationPatterns. A data distribution is a meta representation of individual units of scale-coupling data distributed across a single scale-task. The implementation of the coupling data is arbitrary (so long as it can be serialized for communication). A data distribution represents the smallest unit of data important to the scale-coupling communication, which is necessarily distinct for various multi-scale use cases.

\label{amsi_data_migration_balancing}
If a local installation of the Zoltan \cite{ZoltanOverviewArticle2002} \cite{ZoltanIsorropiaOverview2012} load balancing library is available, the load balancing planning algorithms of Zoltan can be used underlying the scale-sensitive load balancing planning of AMSI. A minimal set of load-balancing algorithms are provided with AMSI. Since there is no general solution to the dynamic load balancing problem any set of provided load-balancing algorithms will prove insufficient for some use case. Thus AMSI provides users with the ability to create and register their own load-balancing algorithms specific to the requirements of a given numerical implementation.

Due to the necessity of retaining inter-scale linkages through the load-balancing processes, the actual data migration undertaken to implement a specific load balancing plan is typically conducted internally by the AMSI system which is provided with buffers of data being load-balanced by each process on the appropriate scale. However, in order to maintain usability in more general situations, it is also possible to simply inform the AMSI system that the current load-balancing plan has already been accomplished by user-implemented data migration. This functionality combined with the ability to use user-designed load-balance planning algorithms allows AMSI to work with entirely external load-balancing libraries and algorithms, allowing scale-sensitive load-balancing to be used even when a specific multi-scale use case falls outside the bounds of capabilities built into the system.

\section{Biotissue Metadata}
